---
title: "ModifyingSSRMetaData"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#clear environment
rm(list = ls())
```

### Load required packages

```{r}
library(dplyr); library(stringr); library(fuzzyjoin); library(lubridate)
```

### Reading in SSR Meta Data

Read in SSR metadata. 

```{r}
SSRM <- read.csv("~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/Metadata/SEW_SSRMetaDataRaw.csv") #this contains all animals
```

Strip white space from all but the CONTENTS column.

```{r}
SSRMetaData1 <- SSRM %>%
                mutate(., INDIVIDUAL = str_trim(INDIVIDUAL)) %>%
                mutate(., SAMPLE_ID = str_trim(SAMPLE_ID)) %>%
                mutate(., CONTENTS = str_trim(CONTENTS))

SSRM <- SSRMetaData1 #whitespace has been removed from the INDIVIDUAL and SAMPLE_ID columns
```

#### Adding rainfall and temperature data

Read in rainfall and temperature data.

```{r}
SSRRainTemp14_16 <- read.csv("~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/SupportingDocs/SSR_2014_2016_RainTemp.csv")
```

Joining rainfall and temperature data to metadata. When left joining, take a moment to check the length of the SSRM df. As a note, due to an equipment malfunction, there is temperature data missing from February 3-8, 2015, May 12-June8, 2015, and March 25-27, 2016. In the left_join below, these entries will be filled with NAs.

```{r}
length(SSRM$SAMPLE_ID) #517 rows

SSRMetaData2 <- SSRM %>%
               left_join(., SSRRainTemp14_16, by = c("DATE" = "DateTime")) #; length(SSRMetaData2$SAMPLE_ID); View(SSRMetaData2) #also 517 rows

SSRM <- SSRMetaData2 #saving the rain/temp additions into the main DF
```

#### Adding dietary variables

Currently in SSRM, there is a CONTENTS column with a string of 4-letter codes that correspond to fruit species, as well as INSE, CATA, and PITH which are broader codes for insects, caterpillars, and pith, respectively. Here, we will create a few diet-related columns.

```{r}
fruitCodes <- read.csv("~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/SupportingDocs/fruitCodes.csv")
fruits<- as.character(fruitCodes$TaxonCodes) #making a list of the 4-letter codes from the parser
nonFruits <- c("INVE", "PITH", "DAMEFL") #making a list of the non-fruit codes

SSRMetaData3 <- SSRM %>%
                mutate(insectsPresent = str_detect(CONTENTS, "INVE")) %>% #created a binary variable for presence/absence of non-caterpillar invertebrates
                mutate(pithPresent = str_detect(CONTENTS, "PITH")) %>% #created a binary variable for presence/absence of pith
                mutate(fruitPresent = str_detect(CONTENTS, paste(fruits, collapse = "|"))) %>% #created a binary variable for presence/absence of fruits
                mutate(flowerPresent = str_detect(CONTENTS, "DAMEFL")) %>% #created a binary variable for presence/absence of flowers
                mutate(caterPresent = str_detect(CONTENTS, "CATA")) %>% #created a binary variable for presence/absence of caterpillars
                mutate(dietType = case_when(fruitPresent == "TRUE" & insectsPresent == "TRUE" ~ "Insect_Fruit", # create general dietary variables 
                                            fruitPresent == "FALSE" & insectsPresent == "TRUE" & caterPresent == "FALSE" ~ "Insect",
                                            fruitPresent == "TRUE" & insectsPresent == "FALSE" & caterPresent == "FALSE" ~ "Fruit",
                                            fruitPresent == "FALSE" & insectsPresent == "FALSE" & caterPresent == "TRUE" ~ "Caterpillar",
                                            fruitPresent == "FALSE" & insectsPresent == "FALSE" & caterPresent == "FALSE" & pithPresent == "TRUE" ~ "Pith",
                                            fruitPresent == "TRUE" & insectsPresent == "FALSE" & caterPresent == "TRUE" ~ "Fruit_Caterpillar",
                                            fruitPresent == "FALSE" & insectsPresent == "FALSE" & caterPresent == "FALSE" & pithPresent == "FALSE" & flowerPresent == "FALSE" ~ "unknown",
                                            fruitPresent == "FALSE" & insectsPresent == "TRUE" & caterPresent == "TRUE" ~ "Insect_Caterpillar"))
SSRM <- SSRMetaData3 #saving new variables in SSRM

SSRMetaData4 <- SSRM %>%
                mutate(dietType = replace(dietType, SampleOrControl == "Control", NA)) %>% #changing dietType of controls to NA
                mutate(IND_CODE = replace(IND_CODE, IND_CODE == "La", "LA")) %>% #fixing a weird capitalization error
                mutate(INDIVIDUAL = replace(INDIVIDUAL, INDIVIDUAL == "legolas", "Legolas")) #fixing capitalization

SSRM <- SSRMetaData4 #saving new changes in SSRM
```

Adding column for colour vision phenotype

```{r}
SSRColourVision <- read.csv("~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/SupportingDocs/2016ColorVision.csv")

SSRMetaData5 <- SSRM %>%
               left_join(., SSRColourVision, by = "INDIVIDUAL") #%>% #adding colour vision phenotype to individuals
               #select(., -c(X.y, X.1, X.2, X.3, X.4, X.5, X.6, X.7, X.8, X.9))

SSRM <- SSRMetaData5

SSRM <- SSRM %>%
        mutate(DATE = as.Date(DATE))
```

### Adding reproductive status to SSRM

For the purposes of the pregnancy microbiome project, we will only be analysing females. Here, we will join reproductive status to metadata.

```{r}
repStatusTable <- read.csv("~/Desktop/PregnancyPaper2019/BehaviouralAnalysis/EcologicalDataCensus/RepStatusSSRFemalesClean.csv")

repStatusTable <- repStatusTable %>%
                  mutate(StartDate = as.Date(StartDate)) %>% #make sure the dates are NOT factors
                  mutate(EndDate = as.Date(EndDate))

SSRMetaData6 <- SSRM %>%
                  fuzzy_left_join(., repStatusTable, 
                            by = c("INDIVIDUAL" = "Animal", "DATE" = "StartDate", "DATE" = "EndDate"), 
                            match_fun = list(`==`, `>=`, `<=`)); View(SSRMetaData6) #adding rep status for all animals; males and females who are not in rep table will have NAs



SSRMetaData8 <- SSRMetaData6 %>%
                dplyr::select(-Animal, -AnimalCode, -X.x, -X.y, -X.1)

SSRM <- SSRMetaData8

write.csv(SSRM,"~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/SEW_SSRMetaData_Clean22April.csv")

SSRM <- read.csv("~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/SEW_SSRMetaData_Clean22April.csv")
```

### Adding binomial variables for ReproductiveStatus

Rename the ReproductiveStatus column (to match the microbiome code). Making additional columns in metadata to make rep status binomial.

In this case, we have categorized Unverified Pregnancies as Pregnant. The commented out code below can be used in cases when UnverifiedPregnancies are treated as UnverifiedPregnancies.

```{r}

SSRM <- rename(SSRM, REP_STATE = ReproductiveStatus) %>% #renaming ReproductiveStatus to REP_STATE
        dplyr::select(-COLLECTOR, -CONSISTENCY, -TemperatureMin, -TIME) #removing these for now

SSRMMetaData9 <- SSRM %>%
                 mutate(Preg.NotPreg = case_when(REP_STATE == "Pregnant" ~ "Pregnant", 
                                                 REP_STATE == "Nursing" ~ "NotPregnant",
                                                 REP_STATE == "Cycling" ~ "NotPregnant")) %>%
                mutate(Nurse.NotNurse = case_when(REP_STATE == "Pregnant" ~ "NotNursing", 
                                                 REP_STATE == "Nursing" ~ "Nursing",
                                                 REP_STATE == "Cycling" ~ "NotNursing")) %>%
                mutate(Cyc.NotCyc = case_when(REP_STATE == "Pregnant" ~ "NotCycling", 
                                                 REP_STATE == "Nursing" ~ "NotCycling",
                                                 REP_STATE == "Cycling" ~ "Cycling")) ; View(SSRMMetaData9)

SSRM <- SSRMMetaData9
```

Filtering unknown reproductive status.

```{r}

SSRMMetaData10 <- SSRM %>%
                  filter(., is.na(REP_STATE) | REP_STATE != "Unknown") %>% #filter will automatically remove NAs so you need to specify that you want them to stay
                  filter(., INDIVIDUAL != "Baloo"); View(SSRMMetaData10) #taking baloo out of the metadata as she had incomplete sequencing

SSRM <- SSRMMetaData10
```

```{r}
write.csv(SSRM,"~/Desktop/PregnancyPaper2019/PregnancyMicrobiomeProject/MicrobiomeAnalysis/Metadata/SSRMetaData16SPregnancy_22April.csv")
```


### Adding stage data for all three states

We decided to run the microbiome analysis on the stages of cycling and pregnancy and lactation. I'm going to read in the prepped meta data, and do an additional join to get the cycling stages in there. 

To prepare the stages of cycling, pregnancy, and lactation:

```{r}
stageDF <- read.csv("RepStatusSSRFemalesCleanCYCLIng_STAGE.csv") #this has cycling stages as well 

SSRM <- read.csv("MicrobiomeAnalysis/Metadata/SEW_SSRMetaData_Clean22April.csv") #reading in clean metadata
```

Left joining the updated rep statuses (which includes the cycling stages)

```{r}
SSRM <- SSRM %>%
        fuzzy_left_join(., stageDF, 
                            by = c("INDIVIDUAL" = "Animal", "DATE" = "StartDate", "DATE" = "EndDate"), 
                            match_fun = list(`==`, `>=`, `<=`))

colnames(SSRM) #need to filter out the duplicated columns

SSRM <- SSRM %>% 
        select(-X, -Animal, -AnimalCode, -StartDate.y, -EndDate.y, -DaysInState, -Miscarriage.y, -InfantSurvive1Year.y, -InfantSurvive3Years.y) %>%
        dplyr::rename(StartDate = StartDate.x, EndDate = EndDate.x, ReproductiveStatus = ReproductiveStatus.x, Miscarriage = Miscarriage.x, InfantSurvive1Year = InfantSurvive1Year.x, InfantSurvive3Years = InfantSurvive3Years.x,  ReproductiveStage = ReproductiveStatus.y)

View(SSRM) #This now has a column for overall state, and another column for stage
```


To add nursing and pregnancy stages to the metadata, we will bring in the full metadata set, and filter for only the categories we are going to analyse. Gestation is considered 157 days in this dataset, as per Carnegie et al. 2011. Nursing is considered 365 days.

Pregnancy Stage 1: 0-53
Pregnancy Stage 2: 54-104
Pregnancy Stage 3: 105-158

Nursing Stage 1: 0-121
Nursing Stage 2: 122-242
Nursing Stage 3: 243-365

```{r}

SSRM$DATE = as.Date(SSRM$DATE)
SSRM$StartDate = as.Date(SSRM$StartDate)
SSRM$EndDate = as.Date(SSRM$EndDate)

SSRM <- SSRM %>%
        mutate(ReproductiveStage = 
               case_when(ReproductiveStage == "Nursing" & DATE >= StartDate & DATE <= StartDate + days(121) ~ "NursingStage1",
                         ReproductiveStage == "Nursing" & DATE >= StartDate + days(122) & DATE <= StartDate + days(242) ~ "NursingStage2",
                         ReproductiveStage == "Nursing" & DATE >= StartDate + days(243) & DATE <= EndDate ~ "NursingStage3",
                         ReproductiveStage == "Pregnant" & DATE >= StartDate & DATE <= StartDate + days(53) ~ "PregnancyStage1",
                         ReproductiveStage == "Pregnant" & DATE >= StartDate + days(54) & DATE <= StartDate + days(104) ~ "PregnancyStage2",
                         ReproductiveStage == "Pregnant" & DATE >= StartDate + days(105) & DATE <= EndDate ~ "PregnancyStage3",
                         ReproductiveStage == "Cycling" ~ "Cycling",
                         ReproductiveStage == "Cycling_PostWeaning" ~ "Cycling_PostWeaning",
                         ReproductiveStage == "Cycling_Pre_Conception" ~ "Cycling_Pre_Conception",
                         ReproductiveStage == "Cycling_PostMiscarriage" ~ "Cycling_PostMiscarriage",
                         ReproductiveStage == "Cycling_PostInfantDeath" ~ "Cycling_PostInfantDeath"))

write.csv(SSRM, "SSRMetaData_AllInd_22July.csv")

SSRM <- filter(SSRM, SEX != "Male")

write.csv(SSRM, "SSRMetaData_FemalesControls_22July.csv" )
```

